### 2014-10-09 paper/include/dtail.tex
write write

Z. Cui et al. \cite{dtail} proposes DTail. They evaluates two versions of DTail referenced as DTail-R and DTail-V utilizing \textbf{R} and \textbf{V} data respecively. Even so, their main idea is that more types of refresh information can be incorperated and stored together with the \textbf{R} and \textbf{V} information. DTail can then choose the most efficient refresh schema depending on the current situation, i.e. workload and type of application. Refresh data is kept per row and is stored in the DRAM itself instead of storing it in the memory controller which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}. To counter the added latency of a slower memory they utilizes spatial locality of the refresh data and performs prefetching into a FIFO buffer. As the refreash data is kept on a row level DTail can save more refreshes than other techniques that have a lower granularity.

@@

When only DTail-R is used, they manages to decrease the amount of refreshes with as much as 87.9\% and thereby only have 16.7\% energy overhead compared to a refresh less 32GB DRAM. With only DTail-V power consumption and performance overhead increase from 4\% and 3\% to 90\% and 23\% respectively when going from 10\% to 90\% memory capacity utilization compared to a refresh less 32 GB system. The power of DTail shows when DTail-R and DTail-V are used together and at 10\% memory utilization a performance overhead of only 1\% and close to 0\% increase in energy consumption is achived. When the utilization is increased to 90\% they increase to 2\% and 15\% respectively.
%

%One thing that DTail-R misses is refresh rate compensation for temperature variations.

>>>>

Z. Cui et al. \cite{dtail} proposes DTail. They evaluates two versions of DTail referenced as DTail-R and DTail-V utilizing \textbf{R} and \textbf{V} data respectively. Even so, their main idea is that more types of refresh information can be incorporated and stored together with the \textbf{R} and \textbf{V} information. DTail can then choose the most efficient refresh schema depending on the current situation, i.e. workload and type of application. Refresh data is kept per row and is stored in the DRAM itself instead of storing it in the memory controller which many other solutions do \cite{raidr}\cite{smartrefresh}\cite{refrint}. To counter the added latency of a slower memory they utilizes spatial locality of the refresh data and performs prefetching into a FIFO buffer. As the refresh data is kept on a row level DTail can save more refreshes than other techniques that have a lower granularity.

@@

When only DTail-R is used, they manages to decrease the amount of refreshes with as much as 87.9\% and thereby only have 16.7\% energy overhead compared to a refresh less 32GB DRAM. With only DTail-V power consumption and performance overhead increase from 4\% and 3\% to 90\% and 23\% respectively when going from 10\% to 90\% memory capacity utilization compared to a refresh less 32 GB system. The power of DTail shows when DTail-R and DTail-V are used together and at 10\% memory utilization a performance overhead of only 1\% and close to 0\% increase in energy consumption is achieved. When the utilization is increased to 90\% they increase to 2\% and 15\% respectively.

### 2014-10-09 paper/include/introduction.tex
write write

\section{Introduction}
\label{sec:int}
Computer systems today are more powerfull and energy effecient than ever before. Battery powered mobile systems are getting more and more common and power consupmtion correspons to a large part of the cost in driving large data centers. This makes energy consumption an increasingly big system design constraint.

@@

There are two main ways to decrease power consumption in DRAM. Minimizing unnecessary operations and scheduling the needed ones smarter. As such, one straightforward way to mitigate DRAM power consumption is to reduce the amount of unnecessary refreshes operations. Several different approaches on how this can be done exists and some of them are surveyed in this paper. The other way to decrease power consupmtion, scheduling of all needed DRAM accesses smarter, will not be examined in any deeper detail in this paper.

>>>>

\section{Introduction}
\label{sec:int}
Computer systems today are more powerful and energy efficient than ever before. Battery powered mobile systems are getting more and more common and power consumption corresponds to a large part of the cost in driving large data centers. This makes energy consumption an increasingly big system design constraint.

@@

There are two main ways to decrease power consumption in DRAM. Minimizing unnecessary operations and scheduling the needed ones smarter. As such, one straightforward way to mitigate DRAM power consumption is to reduce the amount of unnecessary refreshes operations. Several different approaches on how this can be done exists and some of them are surveyed in this paper. The other way to decrease power consumption, scheduling of all needed DRAM accesses smarter, will not be examined in any deeper detail in this paper.

### 2014-10-09 paper/include/raidr.tex
stuff
write write

RAIDR by Jinghui Liu et al. \cite{raidr} divides all memory lines into different groups, called bins, based on the cell with the shortest retention time in the line. In the basic configuration there are two bins, one with a refreshrate of \(64ms - 128ms\) and one with \(128ms - 256ms\). Lines in the first bin gets refreshed every \(64ms\), lines in the second bin every \(128ms\), and lines not in any bin gets refreshed at the default refresh rate which is set to \(256ms\). The implementation is done inside the memory controller and to minimize the storage needed for keeping track of bin members they use Bloom filters. Thanks to this only 1.25 KB storage is needed for a 32 GB DRAM with the default configuration. The storage requirement can however vary depending the amount of bins used. RAIDR also acomodate for temperature variations by scaling all bins refresh rates using a timer. With RAIDR a performance improvment of 4.1\% and a decreas in powerconsumption by 8.3\% is achieved in a system with 32GB DRAM compared to auto-refresh.

>>>>

RAIDR by J. Liu et al. \cite{raidr} divides all memory lines into different groups, called bins, based on the cell with the shortest retention time in the line. In the basic configuration there are two bins, one with a refresh rate of \(64ms - 128ms\) and one with \(128ms - 256ms\). Lines in the first bin gets refreshed every \(64ms\), lines in the second bin every \(128ms\), and lines not in any bin gets refreshed at the default refresh rate which is set to \(256ms\). The implementation is done inside the memory controller and to minimize the storage needed for keeping track of bin members they use Bloom filters. Thanks to this only 1.25 KB storage is needed for a 32 GB DRAM with the default configuration. The storage requirement can however vary depending the amount of bins used. RAIDR also accommodate for temperature variations by scaling all bins refresh rates using a timer. With RAIDR a performance improvement of 4.1\% and a decrease in power consumption by 8.3\% is achieved in a system with 32GB DRAM compared to auto-refresh.

### 2014-10-09 paper/include/reducing_refresh_power.tex
write write

Smart Refresh:
Employed its solution to 2 GB ram. What about todays sizes? How effective is the technique now? - Only works effective if large parts the memory is acccessed, i.e. large part of the memory has to be valid to begin with. Is it so today?

Smart Refresh \& Refrint:
Refrint is somewhat a extension of Smart Refresh. Smart Refresh's features can be laid somewhat equal to Refrint's time-based polices, wheras the data-based polices have no eqvivalence in Smart Refresh.

They both modify the memory controller, and almost in the same way. But Refrint has no timers that is getting decremented the whole time, and instead copies a global "timer" to the line associated bits upon access. Of cource Refrint's data-based polices also adds logic to the memory controller.

>>>>

Smart Refresh:
Employed its solution to 2 GB ram. What about todays sizes? How effective is the technique now? - Only works effective if large parts the memory is accessed, i.e. large part of the memory has to be valid to begin with. Is it so today?

Smart Refresh \& Refrint:
Refrint is somewhat a extension of Smart Refresh. Smart Refresh's features can be laid somewhat equal to Refrint's time-based polices, whereas the data-based polices have no equivalence in Smart Refresh.

They both modify the memory controller, and almost in the same way. But Refrint has no timers that is getting decremented the whole time, and instead copies a global "timer" to the line associated bits upon access. Of course Refrint's data-based polices also adds logic to the memory controller.

### 2014-10-09 paper/include/rioparis.tex
write write

RIO is a technique proposed by Seungjae Baek et al. \cite{rioparis} in which they identify cells with low retention time (bad cells) and `deletes' them. This is done by changing the memory allocating modules in the OS to not allocate pages that have bad cells in them. They implement RIO in a Linux kernel which is run and tested on a real hardware system and is able to achieve 87.5\% reduction in refresh count with a performance increas of 4.5\% on average. As the implementaion only relies on changes in the kernel and there is no need to modify the hardware this technique would be easy to adopt in systems today. To keep fragmentation of the memory low a maximum of 0.1\% of total memory pages gets `deleted'. Even though deletion of a larger amount of pages could decrease the amount of refreshes further it would cause trouble for the kernel which relies on being able to allocate large contiguous address spaces. BLASDLBASDAHSUDGASJHDGJHASGDJHASGDJHAGSDJHGASJHDGAJSHDand it is temperature aware. Seungjae Baek et al. also propose PARIS discussed in \ref{sec:val} which can be combined with RIO for further improvments.

>>>>

RIO is a \textbf{R} based technique proposed by S. Baek et al. \cite{rioparis} in which they identify cells with low retention time (bad cells) and `deletes' them. This is done by changing the memory allocating modules in the OS to not allocate pages that have bad cells in them. They implement RIO in a Linux kernel which is run and tested on a real hardware system and is able to achieve 87.5\% reduction in refresh count with a performance increase of 4.5\% on average. It is shown that going for a refresh period longer than \(256ms\) only yields a small performance and power improvement and therefore they does not push it higher than that. Their implementation almost only relies on changes in the kernel and there is no need to modify the hardware so this technique would be easy to adopt in systems today. To keep fragmentation of the memory low a maximum of 0.1\% of total memory pages gets `deleted'. Even though deletion of a larger amount of pages could decrease the amount of refreshes further it would cause trouble for the kernel which relies on being able to allocate large contiguous address spaces. As retention time varies hugely with temperature RIO continuously measures the system temperature and compensate the refresh period for it.

S. Baek et al. also propose PARIS which can be combined with RIO for further improvements. In PARIS the goal is not to make the refresh period as long as possible but to not refresh unused memory rows at all using \textbf{V} data. This is done by keeping track of all page frames allocated and deallocated by the Linux kernel. The page frame data is translated to DRAM row data and kept in a bitmap with one full/empty bit per row group. A row group consist of rows from several DRAM chips and more rows per group gives a smaller storage overhead for the bitmap, but lowers the refresh granularity. The bitmap is stored in the memory controller and as PARIS utilizes RAS-only refresh this needs to be supported.

As mentioned earlier, RIO and PARIS can be combined which results in PARIS removing unnecessary refreshes and RIO lowering the rate at which the performed refreshes are done. This is similar to what Z. Cui et al. does with DTail-RV. \todo[inline]{compare to DTAIL-RV}
